Automatic discovery of cell types and microcircuitry from neural connectomics

### 2014-06-17 docs/paper.tex
mixing text

\todo{Consistent line thickness}
\todo{on adjacency matrix remove text}

@@

\begin{abstract}
  Connectomics produces massive amounts of data, necessitating new
  analysis methods to discover the biological and computational
  structure. It has long been assumed that discovering neuron types
  and their relation to microcircuitry is crucial to understanding
  neural function. Here we developed a nonparametric Bayesian
  technique that identifies neuron types and microcircuitry patterns
  in connectomics data. It combines the information traditionally used
  by biologists, including connectivity, cell body location and the
  spatial distribution of synapses.  We show that the approach
  recovers known neuron types in the retina and enables predictions of
  connectivity, better than simpler algorithms. It also can reveal
  interesting structure in the nervous system of c. elegans, and
  automatically discovers the structure of a microprocessor.  Our
  approach extracts structural meaning from connectomics, enabling new
  approaches of automatically deriving anatomical insights from these
  emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging techniques \autocite{Morgan2013,Zador2012} promise to
quantify the location of each neuron within a volume of interest,
along with its connections to all other neurons. Far exceeding the
capacity of neuroanatomists to trace small circuits, this leads to
enormous datasets that quantify aspects of the nervous system. The impact of
this kind of high-throughput transition has been seen before -- 
rise of sequencing techniques necessitated the
development of novel computational methods to understand genomic
structure, ushering in an era of bioinformatics as an independent
discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined local connectivity patterns
(microcircuits) has been prominent in many areas, ranging from sensory
(e.g. retina, \autocite{Masland2001} via processing (e.g. neocortex
\autocite{Mountcastle1997}) to movement (e.g. spinal cord)
\autocite{Grillner2005}. Indeed, such repeated patterns also exists in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to automatically back
out underlying microcircuitry.

The discovery of structure is a crucial aspect of network
science. Early approaches focused on global graph properties, such as
the types of scaling present in the network \autocite
{WattsStrogatz1998} .  While this approach leads to an understanding
of global network, more recent work aims at identifying very small-scale
repeat patterns, or “motifs” in networks\autocite{Milo2002}

@@

Here we describe a Bayesian non-parametric model that can discover
circuit structure automatically from connectomics data: the cell
types, their spatial patterns of interconnection, and the locations of
somata and synapses. We primarily focus on the recently-released mouse
retina connectome \autocite{Helmstaedter2013}, but additionally
examine the c. elegans connectome \autocite{White1986}, and then
``connectome'' of a classical microprocessor
\autocite{James2010}. Comparing the cell types discovered by the
algorithms with those obtained manually by human anatomists reveals a
high degree of agreement. We present a scalable approach to infer
microcircuitry from connectomics data.

@@

\section{Results}
We start with the generic assumption that cell types exist and affect
soma depth, distribution of synapses, as well as a cell type and
distance dependent connection probability. From this assumption we
develop a Bayesian algorithm that estimates the underlying cell types.
We start with a model for connectivity, the infinite stochastic block
model (iSBM)\autocite{Kemp2006a,Xu2006}, which has been shown to
meaningfully cluster connection graphs. We extend this approach by
adding distance dependence to model salient aspects of
microcircuitry. We additionally model cell body depth as a mixture of
Gaussians and each cell’s synapse density profile as mixture of
mixtures of Gaussians. For example, retinal ganglion cells may synapse
on nearby, but not far away, amacrine cells, with ganglion cells being
superficial and having a broad distribution of synapses.  This
procedure automatically learns the number of cell types, which cells
belong to which types, along with the the distance and type dependent
connectivity probabilities (see fig~\ref{fig:overview}). We apply our
algorithm to datasets from mouse retina, c-elegans and a historical
microprocessor. Anatomists classify cells based on many features,
giving us a meaningful baseline to compare against.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus started out by
simulating data for which we know the correct structure and comparing
the estimated structure based on the algorithm (see methods) with the
one we used for simulation. We find that the model does a good job of
recovering the correct number of cell types,
(fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  Indeed, it does far better than the
existing infinite stochastic block model that assumes that there is no
distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). It appears that the model behaves the way it should and
allows application to relatively large datasets.
<XXX now add extra features>

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden types in
    synthetic data when incorporating spatial information. a.) The
    iSBM (which only uses connectivity information) over-estimates the
    number of classes as it fails to take distance into account,
    whereas our modeling of the combination of distance and
    connectivity finds close to the true number of classes. b.) ARI, a
    measure of ``correct clustering'', for different true class
    counts, between our model and a traditional iSBM. c.) For
    synthetic data, type discovery based only on connectivity are
    tiny spatially-localized groups -- in
    contrast, our model recovers the true spatial extent of the
    underlying types.  }
\label{fig:synthetic}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome.
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix. c.) connectivity
    diagram showing our clusters, as well as the cell depth and known
    cell type of each cell. Our model shows moderate agreement with
    the cell types traditionally identified by anatomist
    laminar-specific connectivity patterns. \todo{Add more text here}
    d.) Connectivity between our clusters as a function of distance --
    the cluster consisting primarily of retinal ganglion cells (lower
    left-center) exhibits the expected near and far connectivity. e.)
    The spatial distribution of our cell types -- each cell type
    tesselates space.}
\label{fig:mouseretina}
\end{figure}

The retina of the mouse \autocite{Masland2001} is an example where we
should expect to have connectivity patterns that are well approximated
by our generative model. It is known that there are multiple classes
of cells that can be grouped into ganglion cells that transmit
information to the rest of the brain, bipolar cells that connect
between different cells, and amacrine cells that feed into the
ganglion cells. Recent research
\autocite{Helmstaedter2013} has produced a large dataset containing
both the types of cells from orthogonal approaches, and also the
connectivity matrix between all reconstructed cells. We want to know
to which level we can classify neurons into types based on this
connectivity only.

The algorithm took less than 2 hours to converge to a good solution,
dividing neurons into a set of cell types (fig~\ref{fig:mouseretina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability (fig~\ref{fig:mouseretina}b,c,d),
which is well approximated by the model fit. Moreover, each type of
cell is rather isotropically distributed across space
(fig~\ref{fig:mouseretina}e) as should be expected for true cell types. The
algorithm is thus able to extract meaningful structure from the data.

Comparing the results of the algorithm to other information sources
allows evaluating the quality of the type determination. Our types
closely reflect the (anatomist-determined) segmentation of cells into
retinal ganglion, narrow amacrine, medium/wide amacrine, and bipolar
cells (fig~\ref{fig:mouseretina}c, outermost ring). We find that the
types we find tend to reflect the known laminar distribution in the
retina (fig~\ref{fig:mouseretina}c, middle ring), which is exciting as
the algorithm did not actually use this type of information.

@@

<xxx discuss the various algorithms and plus and minus 
   Other model classes
   brielfly explain link prediction accuracy and ARI
   show how it compares to human example	
   the more humanlike the information it gets the more humanlike its classification but many techniques allow good link prediction
>

@@

\todo{spatial extent comparison}

We compared the results of our model to the traditional iSBM to see if
the addition of spatial information was necessary. We find that, as
expected from our synthetic experiments, the iSBM massively
over-segmented the data, finding spatially-localized “communities” of
cells which bore little relation to known types.  The average spatial
extent of the types found by the iSBM was $14.6\pm8.1 \mu m$. Using
the anatomists' ground truth, the mean spatial extent of types was
$34.3\pm7.6 \mu m$, quite close to the $34.5\pm 1.0 \mu m$ discovered
by our algorithm. Modeling the distance-dependence prevents
over-segmentation and thus allows more meaningful estimates.

@@

TODO FIGURE 5: 
We know we want the bar graph with ARI and AUC at the same time:
Conn only
conn and dist
conn and dist and synapse profile and depth
LPCM 
Just cell depth 
human dashed line (for link prediction)

@@

 \item \textbf{Acknowledgments} We thank Josh Vogelstein for discussions and reading of the manuscript, Finale Doshi-Velez for early discussions on the model, and Erica Peterson, and Jonathan Glidden for extensive manuscript review. Funding for compute time was provided by Amazon Web Services ``AWS in Education'' grants. 
\item \textbf{Author Contributions} KK and EJ developed model. EJ derived inference, implemented code, tested, and ran experiments. KK and EJ wrote manuscript text and solicited feedback. 
 \item \textbf{Competing Interests} The authors declare that they have no
competing financial interests.
 \item \textbf{Correspondence} Correspondence and requests for materials
should be addressed to E.J.~(email: jonas@mit.edu).
\end{itemize}

\newpage
\section{Methods }
XXX TODO : add n-way cross validation is not done because it would be a pain
XXX TODO that said overfitting may be shared across all models

@@

% TODO add in the thingie
\begin{equation}
  p(\vec{c}, \eta, \theta | R ) \propto \prod_{i, j} p(R(e_i, e_j) | f(d(e_i, e_j) | \eta_{c_ic_j}), \theta) \prod_{m, n} p(\eta_{mn} | \theta)  p(\theta) p(\vec{c} | \alpha) p(\alpha) p(\theta)
\end{equation}

@@

We chain these three kernels together, and then globally
anneal on the likelihood from a temperature of $T=64$ down to 
$T=1$ over 300 iterations unless otherwise indicated, and
then run the chain for another $100$ iterations.

To pick the MAP, we run up to $100$ simultaneous chains in parallel,
initialized from different random initial points in the state space. 
For visualization we pick the chain with the highest log likelihood,
but for all numerical comparisons (including link probability and ARI)
we use this full collection of posterior samples to estimate
the resulting statistics.

\subsubsection{Mixing of our Markov chains} 
Evaluating whether or not approximate inference methods,
such as MCMC, produce samples which are valid approximations 
of the posterior distribution is an ongoing area of research
in the computational statistics community. Here we validate
mixing in two ways:

\begin{itemize}
\item \textbf{synthetic}: For synthetic datasets of sizes comparable to our
real data size, do we recover known ground truth information
after running our markov chains for the appropriate amount of time? 
\todo{Add figure}
\item \textbf{autocorrelation} For various summary statistics 
of the latent space, after a certain amount of time

\end{itemize}

But importantly, regardless of whether our model over- or under-estimates the exact posterior variance about the network, we find points in the latent variable space that are both predictive \textit{and} parsimonious, largely agreeing with the human anatomists and predicting existing connections.

\subsection {Mouse Retina}
\label{supp:mouseretina}
Dense serial electron microscopy of a $114\mu m \times 80 \mu m $ area
in the mouse retina by \autocite{Helmstaedter2013} yielded a listing
of places where neurons come into contact. There were over 1000 cells
originally, and selected the $950$ for which the location of the soma
could be reconstructed from the provided cell plots (soma locations
were not provided by the study's authors in machine-readable
form). Ultimately this left a matrix between the total synapse-like
contact area between all pairs of 950 cells. Area was thresholded at
$0.5\mu m$, determined by hand, to yield a 950 $\times$ 950 entry
matrix that served as input to our algorithm. We measured the distance
between cells using the reconstructed soma centers, and used the
Logistic-Distance spatial relation. Hyperprior griddings are shown in
supplemental section \ref{supp:hyperpriors}.

@@

\subsection{Source code}

@@

@@

\FloatBarrier

\subsection{Microprocessor Netlist annotation}
\label{supp:mos6502netlist}
This is a matrix consisting of the different clusters of electrical nets for 
the registers, with the colors consistent with figure~\ref{fig:mos6502}.

\includepdf[pages=-]{mos6502.clusters.pdf}

\newpage

\FloatBarrier
\subsection{Detailed learned connectivity for mouse retina}

Here we show the recovered distance-dependent connectivty profile
versus the empirical distance-dependent connectivty for the
retina dataset. The Logistic-distance Poisson likelihood closely
matches the empirical likelihood for most classes.

\begin{figure}[h!]
  \centering 
    \centerline{\includegraphics[width=7in,page=2]{mouseretina/retina.1.2.ld.0.0.data-fixed_20_100-anneal_slow_400.0.latent.pdf}}
  \caption{Learned connectivity vs distance between each of the discovered cell types in the mouse retina. The rows are the source type, and the columns are the destination type. Blue is the empirical plot of the probability of connection vs distance for those two types; red is the learned function}.

\label{fig:supp:mousecompareari}
\end{figure}

\FloatBarrier
\subsection{Model Comparison for Mouse Retina}

We compare the classs found by our model, the traditional infinite
stochastic block model, and a finite stochastic block model with the number
of classes set to the best-found number from our iSRM model ($k=12$).

We see from fig~\ref{fig:supp:irmspatialcomparemouse} that the spatial
variance of our classes closely matches the ground truth, while the
other two models find spatially-localized
classs. Figure~\ref{fig:supp:mousecompareari} shows the ARI of our
clustering, the clustering found by the iSBM,
as well as the result of parametrically varying the traditional stochastic
block model for various values of $k$.

The actual clusterings for iSRM, the iSBM, and $SBM_{k=12}$ are shown
in figures \ref{fig:supp:retinabbclusters},
\ref{fig:supp:retinaldclusters}, and
\ref{fig:supp:retinabbfixedclusters} respectively. Each block is a
cluster, with the true cell type indicated along the x-axis, and the
cell ID written to the left or right.

\begin{figure}[h!]
  \centering 
    \centerline{\includegraphics[width=5in]{mouseretina/spatial_var.pdf}}
  \caption{The distribution of the spatial extent of cell types found by each of
the above methods, and ground truth clusters}
\label{fig:supp:irmspatialcomparemouse}
\end{figure}

\begin{figure}[h!]
  \centering 
    \centerline{\includegraphics[width=5in]{mouseretina/ari_vs_cluster.pdf}}
  \caption{ARI vs cluster number for the iSRM, iSBM, and the SBM with varying K.}
\label{fig:supp:mousecompareari}
\end{figure}

\begin{figure}[h!]
  \centering 
    \centerline{\includegraphics[width=6.0in]{mouseretina/retina.1.2.bb.0.0.data-fixed_20_100-anneal_slow_400.truth_latent.pdf}}
  \caption{The cell IDs and found types (grouped) vs known types (y-axis)  found by the infinite stochastic block model, dramatically overclustering relative to ground truth}
\label{fig:supp:retinabbclusters}
\end{figure}

\begin{figure}[h!]

  \centering 
    \centerline{\includegraphics[width=6.0in]{mouseretina/retina.1.2.ld.0.0.data-fixed_20_100-anneal_slow_400.truth_latent.pdf}}
  \caption{The cell IDs and found types (grouped) vs known types (y-axis) by the infinite spatial-relational model.}
\label{fig:supp:retinaldclusters}
\end{figure}

\begin{figure}[h!]

  \centering 
    \centerline{\includegraphics[width=6.0in]{mouseretina/retina.1.2.bb.0.0.data-fixed_20_12-anneal_slow_fixed_400.truth_latent.pdf}}
  \caption{The cell IDs and found types (grouped) vs known types (y-axis) via the stochastic block model with $k=12$}
\label{fig:supp:retinabbfixedclusters}
\end{figure}

\newpage

\FloatBarrier
\subsection{Hyperprior grids}
\label{supp:hyperpriors}

For the mouse retina Logistic-Distance Bernoulli model, we gridded $\mu^{hp}$ and $\lambda^{hp}$ into 40 $\log_{10}$-spaced points 1.0 and 80. $p_{min} \in [0.001, 0.005, 0.01]$ and $p_{max} \in [0.80, 0.90, 0.95, 0.99]$.

@@

\todo{Concerns about parameter sensitivity}
\todo{How did we pick the best one we often show?}

>>>>

@@

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses.  We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  c. elegans, and automatically discovers the structure of a
  microprocessor.  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural structure.  The impact of this kind of
high-throughput transition has been seen before -- rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in an era of bioinformatics
as an independent discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined, type-dependent local
connectivity patterns (microcircuits) has been prominent in many
areas, ranging from sensory (e.g. retina, \autocite{Masland2001} via
processing (e.g. neocortex \autocite{Mountcastle1997}) to movement
(e.g. spinal cord) \autocite{Grillner2005}. These sorts of repeated
computing patterns are a common feature of computing systems, even
arising in human-made computing circuits. It remains an important
challenge to develop algorithms to use anatomical data,
e.g. connectomics, to automatically back out underlying
microcircuitry.

The discovery of structure is a crucial aspect of network
science. Early approaches focused on global graph properties, such as
the types of scaling present in the network \autocite
{WattsStrogatz1998}.  While this approach leads to an understanding
of the global network, more recent work aims at identifying very small-scale
repeat patterns, or “motifs” in networks\autocite{Milo2002}. \todo{motifs vs type}.

@@

To address these challenges, here we describe a Bayesian
nonparametric model that can discover circuit structure automatically
from connectomics data: the cell types, their spatial patterns of
interconnection, and the locations of somata and synapses. We show
that by incorporating this additional information, our model both
accurately predicts the connection as well as agrees
with human neuroanatomists as to the identification of cell types.

We primarily focus on the recently-released mouse retina connectome
\autocite{Helmstaedter2013}, but additionally examine the c. elegans
connectome \autocite{White1986}, and then ``connectome'' of a
classical microprocessor \autocite{James2010}. Comparing the cell
types discovered by the algorithms with those obtained manually by
human anatomists reveals a high degree of agreement. We present a
scalable approach to infer microcircuitry from connectomics data.

@@

\section{Results}

We start with the generic assumption that cell types exist -- a cell
is a member of a single, unobserved type -- and these types affect
soma depth, distribution of synapses, as well as a cell type and
distance dependent connection probability. From this assumption we
develop a Bayesian algorithm that estimates the underlying cell types.
We start with a model for connectivity, the infinite stochastic block
model (iSBM)\autocite{Kemp2006a,Xu2006}, which has been shown to
meaningfully cluster connection graphs while learning the number of
hidden groups, or types. We extend this approach by adding distance
dependence to model salient aspects of microcircuitry. We additionally
model cell body depth as a mixture of Gaussians and each cell’s
synapse density profile as mixture of mixtures of Gaussians.

For example, retinal ganglion cells may synapse on nearby, but not far
away, amacrine cells, with ganglion cells being superficial and having
a broad distribution of synapses.  This procedure automatically learns
the number of cell types, which cells belong to which types, along
with the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). We apply our algorithm to datasets from mouse
retina, c-elegans and a historical microprocessor. Anatomists classify
cells based on many features, giving us a meaningful baseline to
compare against.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus simulate data for
which we know the correct structure and comparing the estimated
structure based on the algorithm (see methods) with the one we used
for simulation. We find that the model does a good job of recovering
the correct number of cell types, (fig~\ref{fig:synthetic}a), the the
cell identities (fig~\ref{fig:synthetic}b), and the spatial extent of
each type (fig~\ref{fig:synthetic}c).  Indeed, it does far better than
the existing infinite stochastic block model that assumes that there
is no distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). It appears that the model behaves the way it should and
allows application to relatively large datasets.

\begin{figure}
  \centering 
  \centerline{\includegraphics{synthetic.ai}}
  \caption{Correct recovery of true numbers of hidden types in
    synthetic data when incorporating spatial information. a.) The
    infinite stochastic block model (which only uses connectivity
    information) over-estimates the number of classes as it fails to
    take distance into account, whereas our modeling of the
    combination of distance and connectivity finds close to the true
    number of classes. b.) ARI, a measure of ``correct clustering'',
    for different true class counts, between our model and a
    traditional iSBM. c.) For synthetic data, type discovery based
    only on connectivity are tiny spatially-localized groups -- in
    contrast, our model recovers the true spatial extent of the
    underlying types.  }
\label{fig:synthetic}
\end{figure}

@@

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome.
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix. c.) connectivity
    diagram showing our clusters, as well as the cell depth and
    anatomist-labeled cell type of each cell. Our model shows moderate
    agreement with the cell types traditionally identified by
    anatomist laminar-specific connectivity patterns. 
    d.) Connectivity between our clusters as a function
    of distance -- the cluster consisting primarily of retinal
    ganglion cells (lower left-center) exhibits the expected near and
    far connectivity. e.)  The spatial distribution of our cell types
    -- each cell type tesselates space.}
\label{fig:mouseretina}
\end{figure}

The mouse retina \autocite{Masland2001} is a neural circuit which we
expect to have connectivity patterns that are well approximated by our
generative model. It is known that there are multiple classes of cells
that can be broadly grouped into: ganglion cells that transmit
information to the rest of the brain, bipolar cells that connect
between different cells, and amacrine cells that feed into the
ganglion cells. Recent research \autocite{Helmstaedter2013} has
produced a large dataset containing both the types of cells from
orthogonal approaches, and also the connectivity matrix between all
reconstructed cells.

The algorithm took less than 2 hours to perform inference, dividing
neurons into a set of cell types (fig~\ref{fig:mouseretina}c, each
wedge is a type). For each pair of neurons there is a specific
distance dependent connection probability
(fig~\ref{fig:mouseretina}b,c,d), which is well approximated by the
model fit. Moreover, each type of cell is rather isotropically
distributed across space (fig~\ref{fig:mouseretina}e) as should be
expected for true cell types.

Comparing the results of the algorithm to other information sources
allows evaluating the quality of the type determination. Our types
closely reflect the (anatomist-determined) segmentation of cells into
retinal ganglion, narrow amacrine, medium/wide amacrine, and bipolar
cells (fig~\ref{fig:mouseretina}c, outermost ring). We find that the
types we find tend to reflect the known laminar distribution in the
retina (fig~\ref{fig:mouseretina}c, middle ring).

@@

Our fully Bayesian model produces a distribution over probable
clusterings.  Figure~\ref{fig:mouseretina_compare} shows this
posterior distribution as a cell-cell coassignment matrix, sorted to
find maximum block structure. Each large, dark block represents a
collection of cells believed with strong probability to be of the same
type. When we plot (fig~\ref{fig:mouseretina_compare}b) the
anatomist-derived cell types along the left, we can see that each
block consists of a roughly-homogeneous collection of types.

We can thus evaluate our model along three sets of parameters: how
closely does our clustering agree with neuroanatomists' knowledge? 
Given two cells, how accurately can our model predict the link
between them? And how closely does the spatial extent (within a layer)
of our identified types agree with the known neuroanatomists.

We find that given the dataset many techniques allow for good
link-predictive accuracy but poor agreement with neuroanatomists. In
fact, when we fit a version of our model fixing the clusterings given
by neuroanatomists \todo{Add something here}. As we incorporate
additional information beyond connectivity into our model (soma
distance, soma depth, synapse density profile) we more closely
approximate human clustering.

@@

@@

@@

 \item \textbf{Acknowledgments} We thank Josh Vogelstein for discussions and reading of the manuscript, Finale Doshi-Velez for early discussions on the model, and Erica Peterson, and Jonathan Glidden for extensive manuscript review. Funding for compute time was provided by Amazon Web Services ``AWS in Education'' grants. 
\item \textbf{Author Contributions} KK and EJ developed model. EJ derived inference, implemented code, tested, and ran experiments. KK and EJ wrote manuscript text and solicited feedback. 
 \item \textbf{Competing Interests} The authors declare that they have no
competing financial interests.
 \item \textbf{Correspondence} Correspondence and requests for materials
should be addressed to E.J.~(email: jonas@eecs.berkeley.edu).
\end{itemize}

\newpage
\section{Methods }

@@

@@

We chain these three kernels together, and then globally anneal on the
likelihood from a temperature of $T=64$ down to $T=1$ over $900$
iterations unless otherwise indicated, and then run the chain for
another $100$ iterations. We then generate at least $20$ samples, each
taken from the end of a single Markov chain initialized from different
random initial points in the state space. For visualization we pick the
chain with the highest log likelihood, but for all numerical
comparisons (including link probability and ARI) we use this full
collection of samples from the posterior distribution to estimate the
resulting statistics.

\subsection{Link Prediction}
As a proxy for link-prediction accuracy we compute the probability of a link
between two cells using each model, trained fully on the data. While this method
is potentially prone to overfitting, the overfitting will be shared across models
and in fact will preferentially bias in favor of competing models
which over-cluster the data. We use a full collection of posterior samples when
computing the link probability, and then compute the area under the ROC curve for
each.

\subsection{Model Comparison}
We compare our model with a standard network clustering model, the
latent-position clutsering model. This model assumes each cell belongs
to one of K clusters, and each cluster is associated with a
$d-$dimensional Gaussian distribution. The probability of a link is
then a function of the distance between the data points in this
continuous-space. We use \autocite{VBLPCM} a variational
implementation provided in R, parametrically varying the number of
latent dimensions and the number of requested groups.  While this
model provides reasonable link predictive accuracy, the clusterings
dramatically disagree with those from human anatomists.

\subsection{Parameters}

Hierarchical generative models can be sensitive to hyperparameter
settings, thus for most hyperparamters we perform inference. In cases
where we cannot we run separate collections of markov chains at
separate settings and show the results across all pooled
parameters. For the case of the mouse retina data, we consisder
maximum link probability $p_{max} \in \{0.95, 0.9, 0.7\}$, variance
scales for the synapse density profile of $\sigma^2 \in \{0.01, 0.1,
1.0\}$ (of normalized depth), and $K \in \{2, 3\}$ possible synapse
density profile mixture components. For the connectivity-distance-only
model we actually perform inference over both $p_{max}$ and $p_{min}$.

\subsubsection{Mixing of our Markov chains} 
Evaluating whether or not approximate inference methods, such as MCMC,
produce samples which are valid approximations of the posterior
distribution is an ongoing area of research in the computational
statistics community. We use a rough proxy here -- synthetic
likelihood evaluation.  For synthetic datasets of sizes comparable to
our real data size, do we recover known ground truth information after
running our markov chains for the appropriate amount of time?

Figures \ref{fig:mixing:ari} and \ref{fig:mixing:scorevstime} shows
the ARI to ground truth and the total log score as a function of
runtime.  We see dramatic changes in log score initially as we vary
the temperature, stabilizing as runtime progresses, for each
chain. Then we see the characteristic jumps between nearby modes
towards the end of the run, in both log score and ARI.  Importantly,
regardless of whether our model over- or under-estimates the exact
posterior variance about the network, we find points in the latent
variable space that are both predictive \textit{and} parsimonious,
largely agreeing with the human anatomists and predicting existing
connections.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=60mm]{mixing.ari.pdf}}
  \caption{}
\label{fig:mixing:ari}
\end{figure}

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=130mm]{mixing.scorevstime.pdf}}
  \caption{}
\label{fig:mixing:scorevstime}
\end{figure}

\subsection {Mouse Retina}
\label{supp:mouseretina}
Dense serial electron microscopy of a $114\mu m \times 80 \mu m $ area
in the mouse retina by \autocite{Helmstaedter2013} yielded a listing
of places where neurons come into contact. There were over 1000 cells
originally, and selected the $950$ for which the location of the soma
could be reconstructed from the provided cell plots (soma locations
were not provided by the study's authors in machine-readable
form). Ultimately this left a matrix between the total synapse-like
contact area between all pairs of 950 cells. Area was thresholded at
$0.1\mu m$, determined by hand, to yield a 950 $\times$ 950 entry
matrix that served as input to our algorithm. We measured the distance
between cells using the reconstructed soma centers, and used the
Logistic-Distance spatial relation. Hyperprior griddings are shown in
supplemental section \ref{supp:hyperpriors}.

@@

\subsection{Source code and data}

@@

and the content of this paper along with scripts to run
experiments and generate all figures can be found at

\href{http://https://github.com/ericmjonas/netmotifs}{https://github.com/ericmjonas/connect-disco-paper/}

All preprocessed data has been made publicaly available as well.

@@

\FloatBarrier
\subsection{Hyperprior grids and hyperprior inference}
\label{supp:hyperpriors}

For the mouse retina Logistic-Distance Bernoulli model, we gridded
$\mu^{hp}$ and $\lambda^{hp}$ into 40 $\log_{10}$-spaced points 1.0
and 80.

@@
